Boundary terms in the action for Causal Sets

### 2014-10-13 paper_v7.tex
explained bound on contributions outside nbh of sigma in 3.1
added coneplot2.pdf

Consider a set of synchronous or Gaussian Normal Coordinates (GNC), $x^\mu= (t,\mathbf x)$, adapted to $\Sigma$ such that in neighbourhood $U_\Sigma$ of $\Sigma$ the line element is
\be
ds^2 = -dt^2 + h_{ij} (t,\mathbf x) dx^i dx^j.
\ee
In these coordinates, the surface $\Sigma$ corresponds to $t=0$, and the coordinate $t$ measures the proper time elapsed along geodesics whose tangent vector is proportional to the surface normal on $\Sigma$.
%\be\label{GNC_metric}
%g_{\mu\nu} (x)=
%\begin{pmatrix}
 %-1&0 \\
% 0&h_{ij} (x)
%\end{pmatrix}.
%\ee
%In order to find $N_{max}$, the number of maximal points below the surface, one has to use the fact that the points have been sprinkled with a Poisson distribution. 
%If we pick a point $x_0= (0,\mathbf{x})\in \Sigma$ on the surface that has the same spatial coordinates as $x= (t,\mathbf{x})$ then the coordinate time, $t$, is the proper time elapsed along the unique geodesic between $x_0$ and $x$. The volume $V (\Sigma,x)$ then only depends on $t$.~\mb{error?}  
The integrals~\eqref{eq:nmax} and~\eqref{eq:nmin} seem intractable as they stand, since the integration is over the entire causal past/future of the surface and the volume expressions will be complicated in the presence of curvature. However, due to $\Sigma$ being closed compact and $M^+$ and $M^-$ of finite volume, we can always find a subneighbourhood of $U_\Sigma$ such that the contribution to the boundary term from points outside of that neighbourhood tends to zero exponentially quickly as $\rho\rightarrow\infty$.\footnote{To see this, let us restrict ourselves to points to the future of $\Sigma$ and hence to the contributions to $\mathbf F_k$. There will be some $t_{max}>0$ such that all points $(t,\mathbf x)$ with $|t|<t_{max}$ lie in $U_\Sigma$. Choose some $t'<t_{max}$, and consider the subneighbourhood $U'_\Sigma=\left\{(t,\mathbf x)\in U_\Sigma: |t|<t'\right\}$. Then given any $\Lambda\in\mathbb R^+$, there exists, for each $\mathbf x$, a value $\rho_\Lambda(\mathbf x)$ such that $V_\blacktriangle(\mathbf x, t)\rho_\Lambda(\mathbf x)>\Lambda$ for all points outside of $U'_\Sigma$. Since $\Sigma$ is compact, $\rho_\Lambda(\mathbf x)$ achieves its maximum on $\Sigma$ and we can choose a finite constant value $\rho_\Lambda\equiv\max_{\mathbf{x}}\rho_\Lambda(\mathbf x)$ such that $\rho_\Lambda V(\mathbf x, t)>\Lambda$ for all $x\in M^+\setminus U'_\Sigma$. Since $M^+$ has finite volume, the total contribution from those points to $\left<\mathbf P_k\right>$ is then bounded from above by $\rho_\Lambda V(M^+\setminus U'_\Sigma)\frac{\Lambda^k}{k!}e^{-\Lambda}$ for large enough $\Lambda$, a bound that can be made arbitrarily small.} 
Hence, the integration ranges in \eqref{eq:nmax} and \eqref{eq:nmin} can be cut off at finite time $t=\pm\varepsilon$ with $\varepsilon$ arbitrarily small, which allows us to expand the time integration-variable about zero. These assumptions of course impose certain regularity conditions on the surface $\Sigma$, but so long as $\Sigma$ is a closed compact manifold and is everywhere spacelike then all we require is that the radius of extrinsic curvature, $R$, satisfies $R\gg l$. The integrals then simplify to
\begin{gather}\label{eq:nmax_and_eq:nmin}
\begin{aligned}
\left\langle \BF{k}\right\rangle & =\rho \int_{\Sigma}d^{d-1}x\int_{-\varepsilon}^{0}dt\:
h^{\frac{1}{2}}\left (1+
\frac{1}{2}\frac{\dot{h}}{h}t+O (t^2)\right)
 \frac{\left (\rho\: V_\blacktriangledown (p)\right)^k}{k!} e^{-\rho V_\blacktriangledown (t,\mathbf x)}
\\
\left\langle \BP{k}\right\rangle & =\rho \int_{\Sigma}d^{d-1}x\int_{0}^{\varepsilon}dt\:
h^{\frac{1}{2}}\left (1+
\frac{1}{2}\frac{\dot{h}}{h}t+O (t^2)\right) \frac{\left (\rho\: V_\blacktriangle (p)\right)^k}{k!} e^{-\rho V_\blacktriangle (t,\mathbf x)}
\end{aligned}
\end{gather}
where $h:= det\left (h_{ij} (0,\mathbf{x})\right)$ and $\dot{}\: := \frac{\partial}{\partial t}$ and the metric determinant has been in expanded in small $t$.

@@

With the boundaries of the integration region in place, we can now write down the integral explicitly in spherical coordinates:
\be\label{eq:VolumeIntegralSpherical}
V_\blacktriangle (p)=\int_{S^{d-2}}
d\Omega_{d-2}
\int_{0}^{r_{max} (\phi)}r^{d-2}dr
\int_{\frac{1}{2}f (\mathbf{x}_p,\phi)r^2}^{-r+T}
d\overline{t}+O (T^{d+2})
\ee
where $r_{max} (\boldsymbol\phi)$ is the value of the radial coordinate for which the base intersects the top at an angle $\boldsymbol\phi$, as shown in Figure \ref{fig:cone_plot}. To find this value we equate the time coordinates of the top and base surfaces to give 
\be
\frac{1}{2}f (\mathbf{x}_p,\boldsymbol\phi){r_{max}}^2 (\boldsymbol\phi)=-r_{max} (\boldsymbol\phi)+T
\ee 
We then solve this for $r_{max} (\boldsymbol\phi)$ and take the positive solution. The solution can be expanded in $T$ and is simply $r_{max}=T+O (T^2)$, with angular dependent terms contributing at $O (T^2)$. The $O (T^2)$ term will contribute at $O (T^{d+2})$ in the volume integral and so can be ignored. Substituting $r_{max}=T$ into (\ref{eq:VolumeIntegralSpherical}) allows us to evaluate the integral, wherein we find 
\begin{figure}[t]
  \centering
    {\includegraphics[scale=0.5]{coneplot}}
     \caption{A $3$-dimensional representation of the region $\mathcal{X}_p$ in RNCs. The top part of $\partial\mathcal{X}_p$ can be approximated as a flat cone, and the base surface intersects the top part at a radial coordinate, $r_{max} (\boldsymbol\phi)$, which will in general be a function of the angles $\boldsymbol\phi$ (in $3$-dimensions there is one angle $\phi$). This function is found by projecting down from the intersection to the $\overline{t}$ plane. The size of the region inside the top and bottom bounding surfaces is the volume we want to calculate.}
     \label{fig:cone_plot}
\end{figure}
\be\label{eq:VolumeNoK}
V_\blacktriangle (p)
=\frac{V_{d-1}}{d}T^d\left (1-\frac{d}{2 (d+1)}\Gamma^{0}_{\;ij} (p_0)A^{i}_{\;\ibar}A^{j}_{\;\jbar}\delta^{\ibar\jbar}T\right)
+O (T^{d+2})
\ee
where the $\delta^{\ibar\jbar}$ comes from the fact that cross terms ($\ibar\neq \jbar$) vanish under the angular integration. The defining relations for $A^{i}_{\;\ibar}$ can now be rearranged to give $A^{i}_{\;\ibar}A^{j}_{\;\jbar}\delta^{\ibar\jbar}=h^{ij} (p_0)$. In GNCs the extrinsic curvature on the surface is given by
\be\label{eq:K}
K
=g^{\mu\nu }\nabla_{\mu}n_{\nu}
=-\Gamma^{0}_{\;ij}h^{ij}=-\frac{1}{2}\frac{\dot{h}}{h}.
\ee
Substituting this into~\eqref{eq:VolumeNoK} we obtain 
\begin{align}
V_\blacktriangle (T,\mathbf x)
&=\frac{V_{d-1}}{d}T^d\left (1+\frac{d}{2 (d+1)}K (0,\mathbf{x})T\right)
+O (T^{d+2}) \label{eq:TopVolumeWithK}\\
V_\blacktriangledown (-T,\mathbf x)
&=\frac{V_{d-1}}{d}T^d\left (1-\frac{d}{2 (d+1)}K (0,\mathbf{x})T\right)
+O (T^{d+2}), \label{eq:BottomVolumeWithK}
\end{align}
having dropped the subscript $p$. Given the volume expressions in GNCs we now proceed to evaluate the integrals for $\left\langle \BP{k}\right\rangle$ and $\left\langle \BF{k}\right\rangle$.

@@

These objects diverge as $\rho^{1-\frac{1}{d}}$ in the limit. The factor of $\rho$ that must be included in the formula for the surface volume is exactly the inverse of this ($\rho^{\frac{1}{d}-1}$), as can be seen from~\eqref{general_area_sum}, if one substitutes in $l=\rho^{-\frac{1}{d}}$. If we include this factor above, and take the limit, we see that the only remaining terms are proportional to the surface volume. The other terms all have negative powers of $\rho$ and therefore vanish as $\rho\rightarrow\infty$. We can take the mean and limit of $\textbf{A}^{ (d)}$ and substitute in the above expressions to find
\begin{gather}\label{eq:area_final_proof}
\begin{aligned}
\lim_{\rho\rightarrow\infty} & \left\langle \textbf{A}^{ (d)}\right\rangle = \lim_{\rho\rightarrow\infty} \frac{\rho^{\frac{1}{d}-1}}{l_p^{d-1}} b_{d}\left (\sum_m p_m \left\langle\textbf{F}_m\right\rangle  + \sum_n q_n \left\langle\textbf{P}_n\right\rangle\right)
\\
= & \lim_{\rho\rightarrow\infty}\Bigg[ \frac{1}{l_p^{d-1}} \left(\sum_m p_m \frac{\Gamma\left (\frac{1}{d}+m \right)}{m!}  + \sum_n q_n\frac{\Gamma\left (\frac{1}{d}+n \right)}{n!} \right) \int_{\Sigma}d^{d-1}x\: \sqrt{h}+O(\rho^{-\frac{1}{d}})\Bigg]
\\
= & \frac{1}{l_p^{d-1}}\int_{\Sigma}d^{d-1}x\: \sqrt{h}
\end{aligned}
\end{gather}
Where the last line is only true if the vectors, $\vec{p}$ and $\vec{q}$ satisfy~\eqref{area_coefficient_relation}. This concludes the proof of the spatial volume claim,~\eqref{eq:conjecture_for_area}.

>>>>

Consider a set of synchronous or Gaussian Normal Coordinates (GNC), $x^\mu= (t,\mathbf x)$, adapted to $\Sigma$ such that in neighbourhood $U_\Sigma$ of $\Sigma$ the line element is
\be
ds^2 = -dt^2 + h_{ij} (t,\mathbf x) dx^i dx^j.
\ee
In these coordinates, the surface $\Sigma$ corresponds to $t=0$, and the coordinate $t$ measures the proper time elapsed along geodesics whose tangent vector is proportional to the surface normal on $\Sigma$.
%\be\label{GNC_metric}
%g_{\mu\nu} (x)=
%\begin{pmatrix}
 %-1&0 \\
% 0&h_{ij} (x)
%\end{pmatrix}.
%\ee
%In order to find $N_{max}$, the number of maximal points below the surface, one has to use the fact that the points have been sprinkled with a Poisson distribution. 
%If we pick a point $x_0= (0,\mathbf{x})\in \Sigma$ on the surface that has the same spatial coordinates as $x= (t,\mathbf{x})$ then the coordinate time, $t$, is the proper time elapsed along the unique geodesic between $x_0$ and $x$. The volume $V (\Sigma,x)$ then only depends on $t$.~\mb{error?}  
The integrals~\eqref{eq:nmax} and~\eqref{eq:nmin} seem intractable as they stand, since the integration is over the entire causal past/future of the surface and the volume expressions will be complicated in the presence of curvature. However, due to $\Sigma$ being closed compact and $M^+$ and $M^-$ of finite volume, we can always find a subneighbourhood of $U_\Sigma$ such that the contribution to the boundary term from points outside of that neighbourhood tends to zero exponentially quickly as $\rho\rightarrow\infty$.\footnote{To see this, let us restrict ourselves to points to the future of $\Sigma$ and hence to the contributions to $\mathbf F_k$. In the GNC system, there will be some $t_{max}>0$ such that all points $(t,\mathbf x)$ with $|t|<t_{max}$ lie in $U_\Sigma$. Choose some $t'<t_{max}$, and consider the subneighbourhood $U'_\Sigma=\left\{(t,\mathbf x)\in U_\Sigma: |t|<t'\right\}$. Then given any $\Lambda\in\mathbb R^+$, there exists, for each $\mathbf x$, a value $\rho_\Lambda(\mathbf x)$ such that $V_\blacktriangle(\mathbf x, t)\rho_\Lambda(\mathbf x)>\Lambda$ for all points outside of $U'_\Sigma$. Since $\Sigma$ is compact, $\rho_\Lambda(\mathbf x)$ achieves its maximum on $\Sigma$ and we can choose a finite constant value $\rho_\Lambda\equiv\max_{\mathbf{x}}\rho_\Lambda(\mathbf x)$ such that $\rho_\Lambda V(\mathbf x, t)>\Lambda$ for all $x\in M^+\setminus U'_\Sigma$. Since $M^+$ has finite volume, the total contribution from those points to $\left<\mathbf P_k\right>$ is then bounded from above by $\rho_\Lambda V(M^+\setminus U'_\Sigma)\frac{\Lambda^k}{k!}e^{-\Lambda}$ for large enough $\Lambda$, a bound that can be made arbitrarily small.} 
Hence, the integration ranges in \eqref{eq:nmax} and \eqref{eq:nmin} can be cut off at finite time $t=\pm\varepsilon$ with $\varepsilon$ arbitrarily small, which allows us to expand the time integration-variable about zero. These assumptions of course impose certain regularity conditions on the surface $\Sigma$, but so long as $\Sigma$ is a closed compact manifold and is everywhere spacelike then all we require is that the radius of extrinsic curvature, $R$, satisfies $R\gg l$. The integrals then simplify to
\begin{gather}\label{eq:nmax_and_eq:nmin}
\begin{aligned}
\left\langle \BF{k}\right\rangle & =\rho \int_{\Sigma}d^{d-1}x\int_{-\varepsilon}^{0}dt\:
h^{\frac{1}{2}}\left (1+
\frac{1}{2}\frac{\dot{h}}{h}t+O (t^2)\right)
 \frac{\left (\rho\: V_\blacktriangledown (p)\right)^k}{k!} e^{-\rho V_\blacktriangledown (t,\mathbf x)}
\\
\left\langle \BP{k}\right\rangle & =\rho \int_{\Sigma}d^{d-1}x\int_{0}^{\varepsilon}dt\:
h^{\frac{1}{2}}\left (1+
\frac{1}{2}\frac{\dot{h}}{h}t+O (t^2)\right) \frac{\left (\rho\: V_\blacktriangle (p)\right)^k}{k!} e^{-\rho V_\blacktriangle (t,\mathbf x)}
\end{aligned}
\end{gather}
where $h:= det\left (h_{ij} (0,\mathbf{x})\right)$ and $\dot{}\: := \frac{\partial}{\partial t}$ and the metric determinant has been in expanded in small $t$.

@@

With the boundaries of the integration region in place, we can now write down the integral explicitly in spherical coordinates:
\be\label{eq:VolumeIntegralSpherical}
V_\blacktriangle (p)=\int_{S^{d-2}}
d\Omega_{d-2}
\int_{0}^{r_{max} (\phi)}r^{d-2}dr
\int_{\frac{1}{2}f (\mathbf{x}_p,\phi)r^2}^{-r+T}
d\overline{t}+O (T^{d+2})
\ee
where $r_{max} (\boldsymbol\phi)$ is the value of the radial coordinate for which the base intersects the top at an angle $\boldsymbol\phi$, as shown in Figure \ref{fig:cone_plot}. To find this value we equate the time coordinates of the top and base surfaces to give 
\be
\frac{1}{2}f (\mathbf{x}_p,\boldsymbol\phi){r_{max}}^2 (\boldsymbol\phi)=-r_{max} (\boldsymbol\phi)+T
\ee 
We then solve this for $r_{max} (\boldsymbol\phi)$ and take the positive solution. The solution can be expanded in $T$ and is simply $r_{max}=T+O (T^2)$, with angular dependent terms contributing at $O (T^2)$. The $O (T^2)$ term will contribute at $O (T^{d+2})$ in the volume integral and so can be ignored. Substituting $r_{max}=T$ into (\ref{eq:VolumeIntegralSpherical}) allows us to evaluate the integral, wherein we find 
\begin{figure}[t]
  \centering
    {\includegraphics[scale=0.5]{coneplot2}}
     \caption{A $3$-dimensional representation of the region $\mathcal{X}_p$ in RNCs. The top part of $\partial\mathcal{X}_p$ can be approximated as a flat cone, and the base surface intersects the top part at a radial coordinate, $r_{max} (\boldsymbol\phi)$, which will in general be a function of the angles $\boldsymbol\phi$ (in $3$-dimensions there is one angle $\phi$). This function is found by projecting down from the intersection to the $\overline{t}$ plane. The size of the region inside the top and bottom bounding surfaces is the volume we want to calculate.}
     \label{fig:cone_plot}
\end{figure}
\be\label{eq:VolumeNoK}
V_\blacktriangle (p)
=\frac{V_{d-1}}{d}T^d\left (1-\frac{d}{2 (d+1)}\Gamma^{0}_{\;ij} (p_0)A^{i}_{\;\ibar}A^{j}_{\;\jbar}\delta^{\ibar\jbar}T\right)
+O (T^{d+2})
\ee
where the $\delta^{\ibar\jbar}$ comes from the fact that cross terms ($\ibar\neq \jbar$) vanish under the angular integration. The defining relations for $A^{i}_{\;\ibar}$ can now be rearranged to give $A^{i}_{\;\ibar}A^{j}_{\;\jbar}\delta^{\ibar\jbar}=h^{ij} (p_0)$. In GNCs the extrinsic curvature on the surface is given by
\be\label{eq:K}
K
=g^{\mu\nu }\nabla_{\mu}n_{\nu}
=-\Gamma^{0}_{\;ij}h^{ij}=-\frac{1}{2}\frac{\dot{h}}{h}.
\ee
Substituting this into~\eqref{eq:VolumeNoK} we obtain 
\begin{align}
V_\blacktriangle (T,\mathbf x)
&=\frac{V_{d-1}}{d}T^d\left (1+\frac{d}{2 (d+1)}K (0,\mathbf{x})T\right)
+O (T^{d+2}) \label{eq:TopVolumeWithK}\\
V_\blacktriangledown (-T,\mathbf x)
&=\frac{V_{d-1}}{d}T^d\left (1-\frac{d}{2 (d+1)}K (0,\mathbf{x})T\right)
+O (T^{d+2}), \label{eq:BottomVolumeWithK}
\end{align}
having dropped the subscript $p$. Given the volume expressions in GNCs we now proceed to evaluate the integrals for $\left\langle \BP{k}\right\rangle$ and $\left\langle \BF{k}\right\rangle$.

@@

These objects diverge as $\rho^{1-\frac{1}{d}}$ in the limit. The factor of $\rho$ that must be included in the formula for the surface volume is exactly the inverse of this ($\rho^{\frac{1}{d}-1}$), as can be seen from~\eqref{general_area_sum}, if one substitutes in $l=\rho^{-\frac{1}{d}}$. If we include this factor above, and take the limit, we see that the only remaining terms are proportional to the surface volume. The other terms all have negative powers of $\rho$ and therefore vanish as $\rho\rightarrow\infty$. We can take the mean and limit of $\textbf{A}^{ (d)}$ and substitute in the above expressions to find
\begin{gather}\label{eq:area_final_proof}
\begin{aligned}
\lim_{\rho\rightarrow\infty} & \left\langle \textbf{A}^{ (d)}\right\rangle = \lim_{\rho\rightarrow\infty} \frac{\rho^{\frac{1}{d}-1}}{l_p^{d-1}} b_{d}\left (\sum_m p_m \left\langle\textbf{F}_m\right\rangle  + \sum_n q_n \left\langle\textbf{P}_n\right\rangle\right)
\\
= & \lim_{\rho\rightarrow\infty}\Bigg[ \frac{1}{l_p^{d-1}} \left(\sum_m p_m \frac{\Gamma\left (\frac{1}{d}+m \right)}{m!}  + \sum_n q_n\frac{\Gamma\left (\frac{1}{d}+n \right)}{n!} \right) \int_{\Sigma}d^{d-1}x\: \sqrt{h}+O(\rho^{-\frac{1}{d}})\Bigg]
\\
= & \frac{1}{l_p^{d-1}}\int_{\Sigma}d^{d-1}x\: \sqrt{h}
\end{aligned}
\end{gather}
Where the last line is only true if the vectors, $\vec{p}$ and $\vec{q}$ satisfy~\eqref{area_coefficient_relation}. This concludes the proof of the spatial volume claim~\eqref{eq:conjecture_for_area}.
